Modeling and leveraging intuitive theories to improve vaccine attitudes

### 2022-02-09 paper/paper-papaja.Rmd
edits ahead of preprint

author: 
  - name          : "Derek Powell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "4701 W. Thunderbird Road, Glendale, AZ 85306"
    email         : "dmpowell@asu.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Data collection
      - Analysis
      - Writing
  - name          : "Kara Weisman"
    affiliation   : "2"
    role:
      - Conceptualization
      - Data collection
      - Analysis
      - Writing
  - name          : "Ellen M. Markman"
    affiliation   : "3"
    role:
      - Conceptualization
      - Writing

affiliation:
  - id            : "1"
    institution   : "Arizona State University"
  - id            : "2"
    institution   : "University of California, Riverside"
  - id            : "3"
    institution   : "Stanford University"

authornote: |
  This manuscript has not yet been peer-reviewed.

abstract: |
  Much of the richness of human thought is supported by people’s intuitive theories---mental frameworks capturing the perceived structure of the world. But intuitive theories can sometimes contain and reinforce misconceptions, such as misconceptions about vaccine safety that discourage vaccination. We argue that addressing misconceptions requires awareness of the broader conceptual contexts in which they are embedded. Here, we developed a cognitive model of the intuitive theory surrounding vaccination decisions. Using this model, we were able to make accurate predictions about how people's beliefs would be revised in light of educational interventions, design an effective new intervention encouraging vaccination, and understand how these beliefs were affected by real-world events. This approach provides the foundation for richer understandings of intuitive theories and belief revision more broadly.

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

Because intuitive theories are powerful tools for making sense of the world, when these theories are mistaken there can be serious consequences. In this paper, we focus on an important and timely example: misconceptions about vaccine safety.

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over 1000 children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed millions of lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @gershman2019; @lewandowsky.etal2012; @jern.etal2014].

Consider some of the beliefs people who are skeptical of childhood vaccination expressed when we asked them to explain their views in an open-ended qualitative survey (conducted pre-pandemic; *n* = 16). Several participants saw vaccines as "unnatural" and expressed concerns about toxic additives in vaccines, such as heavy metals and mercury. One said, “Vaccines are unnatural additions to the body. The human body should naturally fight off the disease and we should not be controlling how nature plays its role on mankind.” Some doubted whether certain diseases were worth worrying over, while others feared that the contents of vaccines “are too strong and too concentrated for a baby's body to properly handle.” A number of respondents were skeptical of the medical community’s objectivity in testing the safety of vaccines. One respondent wrote that vaccines were “A BIG MONEY MAKING SCAM BY THE BIG PHARMS” (emphasis theirs); another thought that health professionals “are brainwashed by the medical media and the government.” To people holding these beliefs, it could well seem unreasonable to vaccinate a 2-month-old baby against 5 or more diseases at once. Moreover, if skeptics believe that the medical community is unduly influenced by pharmaceutical companies, they might doubt the validity of medical findings that favor these companies’ interests. Altogether, these beliefs might sustain misconceptions about vaccine safety, even in the face of counter-evidence.

The way in which evidence affects any particular belief is unavoidably shaped by other interrelated beliefs. Duhem [-@duhem1954] and Quine [-@quine1951] famously argued that scientific theories necessarily incorporate *auxiliary hypotheses* that influence how observations are interpreted: When an observation is inconsistent with a hypothesis under certain auxiliary assumptions, a scientist cannot know whether to revise the hypothesis or the auxiliary assumptions. Similarly, we argue that intuitive theories incorporate auxiliary beliefs that affect how other beliefs are revised in light of evidence [also see @gershman2019].

@@

2) When two beliefs are related by an intuitive theory, evidence affecting one of those beliefs will also affect the other in accordance with their relationship in the intuitive theory (e.g., evidence for B will increase credence in A, and vice versa); and

3) Thus across individuals, the average change in one belief following evidence affecting another belief will be proportional to the correlation between those beliefs across individuals (evidence for B will increase credence in A in proportion to the correlation between beliefs A and B).

Here, we sought to formalize this understanding of intuitive theories by representing an intuitive theory as a Bayesian network. Originally designed as tools for artificial intelligence [@pearl1988], Bayesian networks are a class of probabilistic graphical models that have been deployed widely across cognitive science [e.g. @gopnik.etal2004; @jern.etal2014; @griffiths.tenenbaum2005]. The heart of a Bayesian Network is a graphical structure ---not unlike a "web of belief"---that represents variables as nodes connected by edges, where those edges represent probabilistic or statistical relations.

@@

Fortunately, the formalism of Bayesian networks provides tools for navigating these challenges. Roughly, when the relationships among beliefs are unknown, we can use the qualitative features of intuitive theories to *infer* the structure of an intuitive theory from observed correlations among beliefs. Structure learning algorithms [see @scutari.denis2021] provide a means for searching through the (super-exponentially large) possible space of Bayesian networks to find the structure most likely to have generated some observed data. These approaches allow us to scale the Bayesian network modelling framework beyond the simple models characteristic of early computational work on intuitive theories [e.g., @griffiths.tenenbaum2005].

@@

In Study 1, we used Bayesian network structure learning algorithms to develop a cognitive model of the relationships among 14 beliefs related to vaccine hesitancy in a large sample of U.S. adults. In Study 2, we replicated the success of an existing intervention emphasizing the dangers of childhood diseases [@horne.etal2015] and used our model to provide a more detailed understanding of *how and why* this intervention increases vaccine intentions and its effects on related beliefs. In Study 3a and 3b, we drew inspiration from our model to create and validate the success of  a novel intervention aimed at debunking concerns about toxic additives. Finally, in Study 4 we tested whether people's beliefs changed coherently following relevant real-world events, by examining beliefs before and after serious outbreaks of measles in the United States in early 2019. Taken together, this work illustrates the promise of understanding belief revision in terms of intuitive theories and the utility of discovering and formalizing intuitive theories via computational models. And, it lays the foundation for empirically-validated, theory-based educational interventions that could encourage people to vaccinate their children and themselves against dangerous diseases like measles and COVID-19.

# The present studies

## Qualitative studies and scale development

We drew on a variety of sources, including academic articles on anti-vaccine skepticism [e.g.,  @dube.etal2015; @larson.etal2015; @salmon.etal2015; @williams2014], anti-vaccine websites, and a qualitative survey with self-identified vaccine skeptics (*n*=16) via Amazon Mechanical Turk (MTurk) to generate a list of beliefs that might influence people’s decisions about whether to vaccinate their children (*vaccine intentions*): *vaccine danger*, *toxic additives in vaccines*, *vaccine effectiveness*, *disease rarity*, *disease severity, infant immune system (IIS): weakness*, *IIS: limited capacity*, *IIS: vaccines strain*, *parental protectiveness*, *parental expertise*, *medical skepticism*, *naturalism*, and *holistic balance*. (See main text for descriptions, and SI for the items included in these scales.) We translated these qualitative insights into 14 psychometrically-valid scales; each scale measured agreement with 4-6 statements, including at least one reverse-coded item, and was highly reliable (observed reliability, Study 1: all Cronbach’s $\alpha \geq .73$). We developed 13 of these scales, and included one preexisting scale [the “holistic balance” subscale from @mcfadden.etal2010].

These 14 scales formed the backbone of Studies 1-4.

## General Methods: Studies 1-4

Studies 1-4 were conducted online via Amazon Mechanical Turk between August 2017 and June 2019. Repeat participation was prevented, with no participants involved in more than one study, except as noted in Study 4. A breakdown of demographics and compensation by study is available in the SI.

All participants responded at least once to the 14 belief scales just described. Each scale was presented on a separate survey page, and participants rated a series of statements on a 7-point scale from “Strongly disagree” (coded as -3) to “Strongly agree” (+3). Scales, and questions within each scale, were presented in a random order for each participant. Attention checks (e.g., “Please select somewhat agree”) were embedded within two to four randomly chosen scales and participants who failed any of these checks were excluded from analyses.

We scored each belief scale by averaging a participant’s responses to each item (after reverse-coding where appropriate), then rescaling these values to fall between 0 and 1. We considered this score to represent the participant’s credence in the belief.

## Study 1: Assessing and modeling intuitive theories of vaccination

Our primary outcome of interest was participants’ intentions to vaccinate their children against diseases like measles, mumps, and rubella (*vaccination intentions*). Drawing on a variety of sources, including research on anti-vaccine skepticism, anti-vaccine websites, and our own qualitative work with vaccine skeptics (see SI), we generated a list of 13 additional beliefs that, though surely not exhaustive, might jointly offer a reasonable means of understanding and predicting vaccine intentions.

These included a variety of claims about vaccines, including beliefs about 1) the overall danger of vaccines (*vaccine danger*); 2) *toxic additives in vaccines*; and 3) *vaccine effectiveness*, how effective vaccines are in preventing disease; as well as a variety of specific claims about childhood diseases like measles, mumps, and rubella, including beliefs about 4) *disease rarity* and 5) *disease severity*. Beyond this, we theorized that participants’ understanding of the infant immune system might be relevant, including beliefs that 6) the infant immune system is weak (*IIS: weakness*); 7) the infant immune system is limited in its capacity and can be easily overwhelmed (*IIS: limited capacity*); and 8) vaccines strain the infant immune system (*IIS: vaccines strain*). We also included general theories about parenting and medicine: 9) general *parental protectiveness*; 10) *parental expertise*, namely the belief that parents usually know more about their children’s health than medical experts; and 11) *medical skepticism*, including concerns about pharmaceutical companies and corruption in the medical community; and two two broad worldviews: 12) *naturalism*, a general preference for natural over artificial things; and 13) *holistic balance*, one important aspect of attitudes toward alternative medicine (e.g., “The body is essentially self-healing and the task of a health care provider is to assist the healing process”; [@mcfadden.etal2010]).

We developed psychometrically robust scales to measure these beliefs, then administered these 14 belief scales to participants and in an online survey with `r results$s1$demo$n_include` participants. There were strong intercorrelations among these 14 beliefs confirming that these beliefs are, in fact, closely related to each other and to intentions to vaccinate children (see SI).

We used these intercorrelations to build a cognitive model of the relations among these beliefs via Bayesian network structure learning techniques.

@@

A total of `r results$s1$demo$n_total` people recruited from Amazon Mechanical Turk (mTurk) participated in this study conducted in August of 2017. Participants were paid \$1.60 for about 8 minutes of their time. `r results$s1$demo$n_exclude` participants (6%) failed at least one of two attention check questions and were excluded from further analysis, leaving a final sample of n = `r results$s1$demo$n_include`. After completing all 14 scales, participants were asked standard demographics questions as well as a handful of questions about whether they were or were expecting to be a parent. They were also debriefed with a brief statement noting that vaccines are safe and pointing them toward a WHO website for further information.

@@

Participants (final *n* = `r results$s1$demo$n_include`) responded to the 14 belief scales, followed by demographics questions and a handful of questions about whether they were or were expecting to be a parent. They were then debriefed with a brief statement noting that vaccines are safe and pointing them toward a WHO website for further information.

@@

The goal of this study was to use these data to conduct a bottom-up search for a Bayesian network that would represent the intuitive theory surrounding vaccination decisions as a network of related beliefs. Based on the observed associations among participants' beliefs about each of theses issues, we sought to infer how these states of affairs (e.g., that vaccines are dangerous, that diseases are severe) are connected in people's intuitive theories---how people understand these states of affairs to cause, imply, or otherwise influence one another.

@@

Having transformed participants’ averaged belief scale responses to fall between 0 and 1, we estimated the parameters of the model as a beta regression using maximum likelihood methods (see SI).

@@

We aimed to represent not only the relationships among variables in the data but the *cognitive model* that gave rise to these observed relationships, i.e., the relationships among beliefs in the mind. We actually estimated two models: First, we estimated a data model assuming that participants possess a shared mental model from which they generate their responses to the belief scales. In this model, the $x$ in the equation above corresponded to participants’ continuous responses to the belief scales (e.g. a participants score for the *vaccine danger* scale). Then, from the parameters estimated in the data model we created a cognitive model of participants’ reasoning, where the $x$ are binary, indicating whether that state of affairs obtains (e.g., whether “vaccines are dangerous”). We used this cognitive model to simulate how beliefs could change following exposure to evidence.

@@

In particular, we assumed  that the intuitive theory is a generative model, so that these influence relations (“edges”) should flow from causes or other “generating” states of affairs toward effects. For instance, we stipulated that *vaccine intentions* would be a “child node,” with edges only flowing inwards and not outwards (since one’s intention to vaccinate cannot “generate” the facts about vaccinations safety, effectiveness, etc).

We applied a constraint we called the “generative model principle,” specifying that influence relations (edges) should flow from causes (or other generating states of affairs) to effects; instilling a core assumption in research on intuitive theories and causal explanatory reasoning [e.g. @gerstenberg.etal2021; @gopnik.etal2004; @cheng1997]. We sorted the 14 measured beliefs into four tiers based on how general or abstract each belief was, reasoning that general and abstract states of affairs should be understood to cause (or otherwise generate) more specific and concrete states of affairs, so edges should flow from the general and abstract toward the specific and concrete (e.g., if natural things are better than artificial things, this should set an expectation about whether a specific class of artificial things--vaccines--is safe). We considered *holistic balance* and *naturalism* to be the most general and abstract beliefs among those we measured; followed by theories about parenting and medicine; then specific beliefs about vaccines, diseases, and the infant immune system; and finally our outcome of interest, *vaccine intentions*---the most concrete measurement of a specific intended action.

@@

Some caution should be exercised in interpreting the relationships in the network. Due to Bayesian networks' tight connections with structural causal networks, it is tempting to interpret the edges in this network as representing a mental causal model. But, in some cases these relationships are better thought of as probabilistic dependencies or “influence” relations [@williamson2001]. For instance, intuitively we might imagine that vaccines having toxins could cause them to be dangerous and ineffective. These nodes are connected in the model, but the edges between them run in the opposite directions, and therefore do not support a clear causal interpretation. One interpretation is that this edge instead reflects a set membership relation; that toxins are one type of danger vaccines might present. In other cases, our data and explicit constraints do not perfectly identify the directions of the edges. This is the case for two edges (Naturalism $\rightarrow$  Holistic Balance and Medical Skepticism $\rightarrow$ Parental Expertise; see SI for details). In both cases, we do not feel it is intuitive which direction the edge should flow in, nor even that it should be directed at all; we might even speculate there is an additional unmeasured variable that connects these beliefs and that would best account for their association.

Bearing these qualifications in mind, we see this Bayesian network as a limited but plausible cognitive model of these beliefs.

@@

The three nodes with direct connections to *vaccine intentions* are beliefs about *vaccine effectiveness*, *vaccine danger*, and the severity of childhood diseases (*disease severity*) ---concrete beliefs that are, at face value, especially closely related to vaccination decisions.  This gives us confidence that the model is capturing important relationships.

@@

This model structure also helps explain some initially surprising findings. For example, we originally speculated that dispelling misconceptions about the capacity of the infant immune system could promote positive attitudes toward vaccination. We were disappointed to observe the weak first-order correlation between this belief and *vaccine intentions* in our behavioral data (r = -.09 in our training split). The model sheds light on this surprising (lack of) relationship: Although the belief that the infant immune system is limited in capacity is positively related to the belief that vaccines strain the immune system---discouraging vaccination, as we had assumed---it also seems to promote the belief that childhood diseases have severe consequences for young children (*disease severity*), which might, in turn, encourage vaccination. These countervailing forces---which, on average, seem to have cancelled each other out in this sample---dissuaded us from our plan to develop an intervention aimed at dispelling misconceptions about limited capacity in order to encourage childhood vaccination.

In sum: From a set of potentially relevant beliefs and a general principle guiding the likely relationships among these beliefs, we generated a cognitive model of an intuitive theory of this domain. In Studies 2-4, we use this same model to predict how this intuitive theory guides reasoning about new information.

@@

One of the few successful educational interventions in this domain was tested by Horne et al. [-@horne.etal2015], who found that emphasizing the dangers of preventable diseases like mumps, measles, and rubella improved participants’ attitudes toward vaccines. Their intervention included descriptions of the symptoms of mumps, measles, and rubella; vivid photos of children suffering from these diseases; and a brief story from the perspective of a parent whose child was hospitalized with measles (see SI). This “disease risk” intervention improved attitudes toward vaccines compared to an irrelevant control essay and compared to an intervention aimed at debunking concerns about vaccines causing autism.

>>>>

author: 
  - name          : "Derek Powell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "4701 W. Thunderbird Road, Glendale, AZ 85306"
    email         : "dmpowell@asu.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Investigation
      - Methodology
      - Data collection
      - Formal Analysis
      - Visualization
      - Writing
  - name          : "Kara Weisman"
    affiliation   : "2"
    role:
      - Conceptualization 6t            
      - Investigation
      - Data collection
      - Formal Analysis
      - Visualization
      - Writing
  - name          : "Ellen M. Markman"
    affiliation   : "3"
    role:
      - Conceptualization
      - Methodology
      - Writing

affiliation:
  - id            : "1"
    institution   : "Arizona State University"
  - id            : "2"
    institution   : "University of California - Riverside"
  - id            : "3"
    institution   : "Stanford University"

note: |
  This manuscript has not yet been peer-reviewed.

authornote: |
  All data, and analysis code for this project are available at https://osf.io/sdfpj/. Preregistration information for Studies 3a and 3b is available at https://osf.io/csnez/registrations.

abstract: |
  Much of the richness of human thought is supported by people’s intuitive theories---mental frameworks capturing the perceived structure of the world. But intuitive theories can also contain and reinforce dangerous misconceptions. In this paper we take up the case of misconceptions about vaccine safety that discourage vaccination. These misconceptions constitute a major public health risk that predates the coronavirus pandemic but that has become all the more dire in recent years. We argue that addressing such misconceptions requires awareness of the broader conceptual contexts in which they are embedded. To build this understanding, we examined the structure and revision of people’s intuitive theories of vaccination in five large survey studies (total _N_ = 3196). Based on these data, we present a cognitive model of the intuitive theory surrounding people’s decisions about whether to vaccinate young children against diseases like measles, mumps, and rubella (MMR). Using this model, we were able to make accurate predictions about how people’s beliefs would be revised in light of educational interventions, design an effective new intervention encouraging vaccination, and understand how these beliefs were affected by real-world events (the measles outbreaks of 2019). In addition to presenting a promising way forward for promoting the MMR vaccine, this approach has clear implications for encouraging the uptake of COVID-19 vaccines, especially among parents of young children. At the same time, this work provides the foundation for richer understandings of intuitive theories and belief revision more broadly.

@@

documentclass     : "apa7"
csl               : "apa7.csl"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

Because intuitive theories are such powerful tools for making sense of the world, when these theories are mistaken there can be serious consequences. In this paper, we focus on an important and timely example: misconceptions about vaccine safety.

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over 1000 children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: widespread reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed more than five million lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @gershman2019; @lewandowsky.etal2012; @jern.etal2014]. 
Consider some of the beliefs people who are skeptical of childhood vaccination expressed when we asked them to explain their views in an open-ended qualitative survey (conducted pre-pandemic; _n_ = 16). Several participants saw vaccines as “unnatural” and expressed concerns about toxic additives in vaccines, such as heavy metals and mercury. One said, “Vaccines are unnatural additions to the body. The human body should naturally fight off the disease and we should not be controlling how nature plays its role on mankind.” Some doubted whether certain diseases were worth worrying over, while others feared that the contents of vaccines “are too strong and too concentrated for a baby’s body to properly handle.” A number of respondents were skeptical of the medical community’s objectivity in testing the safety of vaccines. One respondent wrote that vaccines were “A BIG MONEY MAKING SCAM BY THE BIG PHARMS” (emphasis theirs); another thought that health professionals “are brainwashed by the medical media and the government.”

To people holding these beliefs, it could well seem unreasonable to vaccinate a 2-month-old baby against 5 or more diseases at once, as the CDC recommends. Moreover, if skeptics believe that the medical community is unduly influenced by pharmaceutical companies, they might doubt the validity of medical findings that appear to favor these companies’ interests. Altogether, these beliefs might sustain misconceptions about vaccine safety, even in the face of counter-evidence: CDC guidelines, clinical studies, even the advice of the family’s pediatrician could be interpreted not as evidence that vaccines are safer than the person thought, but instead as further proof of the widespread corruption in the medical community.

The way in which evidence affects any particular belief is unavoidably shaped by other interrelated beliefs. Duhem [-@duhem1954] and Quine [-@quine1951] famously argued that scientific theories necessarily incorporate *auxiliary hypotheses* that influence how observations are interpreted: when an observation is inconsistent with a hypothesis under certain auxiliary assumptions, a scientist cannot know whether to revise the hypothesis or the auxiliary assumptions. Similarly, we argue that intuitive theories incorporate auxiliary beliefs that affect how other beliefs are revised in light of evidence: when an observation is inconsistent with a person’s current belief under certain auxiliary assumptions, the person cannot know whether to revise the belief or the auxiliary assumptions [also see @gershman2019].

@@

2) When two beliefs are related by an intuitive theory, evidence affecting one of those beliefs will also affect the other in accordance with their relationship in the intuitive theory (e.g., evidence for B will increase credence in A, and vice versa).

3) Thus, across individuals, the average change in one belief following evidence affecting another belief will be proportional to the correlation between those beliefs across individuals (evidence for B will increase credence in A in proportion to the correlation between beliefs A and B).

Here, we sought to formalize this understanding of intuitive theories by representing an intuitive theory as a Bayesian network. Originally designed as tools for artificial intelligence [@pearl1988], Bayesian networks are a class of probabilistic graphical models that have been deployed widely across cognitive science [e.g. @gopnik.etal2004; @jern.etal2014; @griffiths.tenenbaum2005]. The heart of a Bayesian network is a graphical structure ---not unlike a "web of belief"---that represents variables as nodes connected by edges, where those edges represent probabilistic or statistical relations.

@@

Fortunately, the formalism of Bayesian networks provides tools for navigating these challenges. Roughly, when the relationships among beliefs are unknown, we can use the qualitative features of intuitive theories to *infer* the structure of an intuitive theory from observed correlations among beliefs. Structure learning algorithms [see @scutari.denis2021] provide a means for searching through the (super-exponentially large) possible space of Bayesian networks to find the structure most likely to have generated some observed data. These approaches allow us to scale the Bayesian network modeling framework beyond the smaller models characteristic of early computational work on intuitive theories [e.g., @griffiths.tenenbaum2005].

## Current studies

@@

In Study 1, we used Bayesian network structure learning algorithms to develop a cognitive model of the relationships among 14 beliefs related to vaccine hesitancy in a large sample of U.S. adults. In Study 2, we replicated the success of an existing intervention emphasizing the dangers of childhood diseases [@horne.etal2015] and used our model to provide a more detailed understanding of *how and why* this intervention increases vaccine intentions and its effects on related beliefs. In Study 3a and 3b, we drew inspiration from our model to create and validate the success of  a novel intervention aimed at addressing concerns about toxic additives in vaccines. Finally, in Study 4 we tested whether people's beliefs changed coherently following relevant real-world events, by examining beliefs before and after serious outbreaks of measles in the United States in early 2019.

Taken together, this work illustrates the promise of understanding belief revision in terms of intuitive theories and the utility of discovering and formalizing intuitive theories via computational models. Of particular interest in the current moment, these findings lay a foundation for empirically-validated, theory-based educational interventions that could encourage people to vaccinate their children and themselves against dangerous diseases like measles and COVID-19.

# Study 1: Assessing and modeling intuitive theories of vaccination

Across all our studies, our primary outcome of interest was participants’ intentions to vaccinate their children against diseases like measles, mumps, and rubella (which we refer to henceforth as *vaccination intentions*). Drawing on a variety of sources, including research on anti-vaccine skepticism [e.g.,  @dube.etal2015; @larson.etal2015; @salmon.etal2015; @williams2014], anti-vaccine websites, and our own qualitative work with vaccine skeptics (see SI), we generated a list of 13 additional beliefs that, though surely not exhaustive, might jointly offer a reasonable means of understanding and predicting vaccine intentions.

We translated these qualitative insights into 14 psychometrically-valid scales. In addition to 1) _vaccination intentions_ (e.g. “If I had a young baby I would have him or her vaccinated.”), these included a variety of beliefs about vaccines, including beliefs about 2) _vaccine danger_, the overall danger of vaccines (e.g., “Many children have severe adverse reactions to the MMR vaccine”); 3) _toxic additives in vaccines_ (e.g., “Exposure to certain additives in vaccines can cause major health problems.”); and 4) _vaccine effectiveness_, how effective vaccines are in preventing disease (e.g., “Your chances of getting a disease after being vaccinated against it are incredibly low”); as well as a variety of specific claims about childhood diseases like measles, mumps, and rubella, including beliefs about 5) _disease rarity _(e.g., “Diseases like measles and whooping cough are so rare that there is no real need to worry about them”) and 6) _disease severity_ (e.g., “Diseases like measles and whooping cough are extremely dangerous for young children”). Beyond this, we theorized that participants’ understanding of the infant immune system might be relevant, including beliefs that 7) the infant immune system is weak (_IIS: weakness_; e.g., “Babies’ immune systems are weak”); 8) the infant immune system is limited in its capacity and can be easily overwhelmed (_IIS: limited capacity_; e.g., “Babies’ immune systems are severely limited in how much they can cope with at one time.”); and 9) vaccines strain the infant immune system (_IIS: vaccines strain_; e.g., “Vaccines strain the immune system much like the actual disease”). We also included general theories about parenting and medicine: 10) general _parental protectiveness_ (e.g., “Parents should strive to control their baby’s environments as much as possible”); 11) _parental expertise_, namely the belief that parents usually know more about their children’s health than medical experts (e.g., “Every individual is different and a parent knows their own child best”); and 12) _medical skepticism_, including concerns about pharmaceutical companies and corruption in the medical community (e.g., “Pharmaceutical companies put pressure on the FDA and CDC to suppress negative findings”); and two two broad worldviews: 13) _naturalism_, a general preference for natural over artificial things (e.g., “Natural things are always better than synthetic alternatives”); and 14) _holistic balance_, one important aspect of attitudes toward alternative medicine (e.g., “The body is essentially self-healing and the task of a health care provider is to assist the healing process”; [@mcfadden.etal2010]).

Each scale measured agreement with 4-6 statements, including at least one reverse-coded item, and was highly reliable (observed reliability, Study 1: all Cronbach’s $\alpha \geq .73$ .73). We developed 13 of these scales, and included one preexisting scale (the “holistic balance” subscale from [@mcfadden.etal2010]. See SI for all items included in all scales.

In Study 1, we administered these 14 belief scales to participants and in an online survey with `r results$s1$demo$n_include` participants. We then used these intercorrelations to build a cognitive model of the relations among these beliefs via Bayesian network structure learning techniques.

@@

A total of `r results$s1$demo$n_total` people recruited from Amazon Mechanical Turk (mTurk) participated in this study conducted in August of 2017. Participants were paid \$1.60 for about 8 minutes of their time. `r results$s1$demo$n_exclude` participants (6%) failed at least one of two attention check questions and were excluded from further analysis, leaving a final sample of n = `r results$s1$demo$n_include`. See SI for demographic information about this sample.

@@

Participants (final *n* = `r results$s1$demo$n_include`) responded to the 14 belief scales. Each scale was presented on a separate survey page, and participants rated a series of statements on a 7-point scale from “Strongly disagree” (coded as -3) to “Strongly agree” (+3). Scales, and questions within each scale, were presented in a random order for each participant. Attention checks (e.g., “Please select somewhat agree”) were embedded within two to four randomly chosen scales and participants who failed any of these checks were excluded from analyses. We scored each belief scale by averaging a participant’s responses to each item (after reverse-coding where appropriate), then rescaling these values to fall between 0 and 1. We considered this score to represent the participant’s credence in the belief.

The 14 belief scales were followed by demographics questions and a handful of questions about whether participants were or were expecting to be a parent. Participants were then debriefed with a brief statement noting that vaccines are safe and pointing them toward a WHO website for further information.

@@

The goal of this study was to use these data to conduct a bottom-up search for a Bayesian network that would represent the intuitive theory surrounding vaccination decisions as a network of related beliefs. There were strong intercorrelations among these 14 beliefs confirming that these beliefs are, in fact, closely related to each other and to intentions to vaccinate children (see SI). Based on the observed associations among participants' beliefs about each of theses issues, we sought to infer how these states of affairs (e.g., that vaccines are dangerous, that diseases are severe) are connected in people's intuitive theories---how people understand these states of affairs to cause, imply, or otherwise influence one another.

@@

Having transformed participants’ scores on each belief scale to fall between 0 and 1, we estimated the parameters of the model as a beta regression using maximum likelihood methods (see SI).

@@

We aimed to represent not only the relationships among variables in the data but the *cognitive model* that gave rise to these observed relationships, i.e., the relationships among beliefs in the mind. We actually estimated two models: First, we estimated a data model assuming that participants possess a shared mental model from which they generate their responses to the belief scales. In this model, the $x$ in the equation above corresponded to participants’ continuous responses to the belief scales (e.g. a participant's score for the *vaccine danger* scale). Then, from the parameters estimated in the data model we created a cognitive model of participants’ reasoning, where the $x$ are binary, indicating whether that state of affairs obtains (e.g., whether “vaccines are dangerous”). We used this cognitive model to simulate how beliefs could change following exposure to evidence.

@@

In particular, we applied a constraint we called the “generative model principle,” based on the assumption that the intuitive theory is a generative model, so that these influence relations (“edges”) should flow from causes or other “generating” states of affairs toward effects. For instance, we stipulated that vaccine intentions---our outcome of interest---would be a “child node,” with edges only flowing inwards and not outwards (since one’s intention to vaccinate cannot “generate” the facts about vaccinations safety, effectiveness, etc). This generative model principle instills a common core assumption in research on intuitive theories and causal explanatory reasoning[e.g. @gerstenberg.etal2021; @gopnik.etal2004; @cheng1997].

To apply our generative model principle, we sorted the 14 measured beliefs into four tiers based on how general or abstract each belief was, reasoning that general and abstract states of affairs should be understood to cause (or otherwise generate) more specific and concrete states of affairs, so edges should flow from the general and abstract toward the specific and concrete (e.g., if natural things are better than artificial things, this should set an expectation about whether a specific class of artificial things--vaccines--is safe). We considered *holistic balance* and *naturalism* to be the most general and abstract beliefs among those we measured; followed by theories about parenting and medicine; then specific beliefs about vaccines, diseases, and the infant immune system; and finally our outcome of interest, *vaccine intentions*---the most concrete measurement of a specific intended action.

@@

@@

The three nodes with direct connections to *vaccine intentions* are beliefs about *vaccine effectiveness*, *vaccine danger*, and the severity of childhood diseases (*disease severity*) ---concrete beliefs that are, at face value, especially closely related to vaccination decisions.  This gives us confidence that the model is capturing important aspects of the intuitive theory of interest.

@@

This model structure also helps explain some initially surprising findings. For example, we originally speculated that dispelling misconceptions about the capacity of the infant immune system could promote positive attitudes toward vaccination. We were disappointed to observe the weak first-order correlation between this belief and *vaccine intentions* in our behavioral data (r = -.09 in our training split). The model sheds light on this surprising (lack of) relationship: Although the belief that the infant immune system is limited in capacity is positively related to the belief that vaccines strain the immune system---discouraging vaccination, as we had assumed---it also seems to promote the belief that childhood diseases have severe consequences for young children (*disease severity*), which might, in turn, encourage vaccination. These countervailing forces---which, on average, seem to have cancelled one-another out in this sample---dissuaded us from our plan to develop an intervention aimed at dispelling misconceptions about limited capacity in order to encourage childhood vaccination.

It is worth pointing out that not all of the relationships represented in the model should be interpreted as _causal_. Due to Bayesian networks’ tight connections with structural causal networks, it can be tempting to give the edges this interpretation, but in some cases these relationships are better thought of as probabilistic dependencies or “influence” relations (Williamson, 2001). For instance, intuitively we might imagine that vaccines having toxins (_toxic additives in vaccines_) could cause them to be dangerous (_vaccine danger_). These nodes are connected in the model, but the edges between them run in the opposite direction, and therefore do not support a clear causal interpretation. One interpretation is that this edge instead reflects a set membership relation; that toxins are one type of danger vaccines might present.

In other cases, our data and explicit constraints do not perfectly identify the directions of the edges. This is the case for two edges (Naturalism $\rightarrow$  Holistic Balance and Medical Skepticism $\rightarrow$ Parental Expertise; see SI for details). In both cases, we do not feel it is intuitive which direction the edge should flow in, nor even that it should be directed at all; we might even speculate there is an additional unmeasured variable that connects these beliefs and that would best account for their association.

### Summary

From a set of potentially relevant beliefs and a general principle guiding the likely relationships among these beliefs, we generated a cognitive model of an intuitive theory of this domain. In Studies 2-4, we use this same model to predict how this intuitive theory guides reasoning about new information.

@@

Very few educational interventions have been shown to improve attitudes toward childhood vaccination. One of the few successful educational interventions in this domain was tested by Horne et al. [-@horne.etal2015], who found that emphasizing the dangers of preventable diseases like mumps, measles, and rubella improved participants’ attitudes toward vaccines. Their intervention included descriptions of the symptoms of mumps, measles, and rubella; vivid photos of children suffering from these diseases; and a brief story from the perspective of a parent whose child was hospitalized with measles (see SI). This “disease risk” intervention improved attitudes toward vaccines compared to an irrelevant control essay and compared to an intervention aimed at debunking concerns about vaccines causing autism. In Study 2, we aimed to replicate this success and leverage the cognitive model developed in Study 1 to gain insight into how and why such an intervention successfully induces  belief revision in this domain.
